---
title: "HW 3"
author: "Swagat Adhikary"
date: "9/24/2024"
output: 
  html_document:
    number_sections: true
---

#

Let $E[X] = \mu$.  Show that $Var[X] := E[(X-E[X])^2] = E[X^2]-(E[X])^2$.  Note, all you have to do is show the second equality (the first is our definition from class). 

\[
\begin{align*}
Var[X] & = E[(X-E[X])^2] \\
       & = E[X^2 - 2X \cdot E[X] + (E[X])^2] \\
       & = E[X^2] - 2E[X \cdot E[X]] + E[(E[X])^2] \\
       & = E[X^2] - 2 \cdot E[X] \cdot E[X] + (E[X])^2 \quad \text{(since } E[X] \text{ is a constant)} \\
       & = E[X^2] - 2(E[X])^2 + (E[X])^2 \\
       & = E[X^2] - (E[X])^2
\end{align*}
\]

# 

In the computational section of this homework, we will discuss support vector machines and tree-based methods.  I will begin by simulating some data for you to use with SVM. 

```{r}
library(e1071)
set.seed(1) 
x=matrix(rnorm(200*2),ncol=2)
x[1:100,]=x[1:100,]+2
x[101:150,]=x[101:150,]-2
y=c(rep(1,150),rep(2,50))
dat=data.frame(x=x,y=as.factor(y))
plot(x, col=y)

```


##

Quite clearly, the above data is not linearly separable.  Create a training-testing partition with 100 random observations in the training partition.  Fit an svm on this training data using the radial kernel, and tuning parameters $\gamma=1$, cost $=1$.  Plot the svm on the training data.  

```{r}
set.seed(1)
train=sample(200,100)
svmfit = svm(y ~ ., data = dat[train,], kernel = "radial", gamma = 1, cost = 1)
plot(svmfit, dat[train,])

```

##

Notice that the above decision boundary is decidedly non-linear.  It seems to perform reasonably well, but there are indeed some misclassifications.  Let's see if increasing the cost ^[Remember this is a parameter that decides how smooth your decision boundary should be] helps our classification error rate.  Refit the svm with the radial kernel, $\gamma=1$, and a cost of 10000.  Plot this svm on the training data. 

```{r}
svmfit = svm(y ~ ., data = dat[train,], kernel = "radial", gamma = 1, cost = 10000)
plot(svmfit, dat[train,])

```

##

It would appear that we are better capturing the training data, but comment on the dangers (if any exist), of such a model. 

*While increasing the cost parameter allows the model to fit the training data more accurately, it can also lead to overfitting. A high-cost SVM has a more complex decision boundary that classifies the training data very well but might perform poorly on unseen data due to its inability to generalize well.*

##

Create a confusion matrix by using this svm to predict on the current testing partition.  Comment on the confusion matrix.  Is there any disparity in our classification results?    

```{r}
table(true=dat[-train,"y"], pred=predict(svmfit, newdata=dat[-train,]))

```


##

Is this disparity because of imbalance in the training/testing partition?  Find the proportion of class `2` in your training partition and see if it is broadly representative of the underlying 25\% of class 2 in the data as a whole.  

```{r}
sum(dat[train,3] == 2)/100

```

*The proportion of class 2 in my training set is 29% which is pretty similar to the underlying 25%. Therefore, this disparity is not due to an imbalance in the training/testing partition and seems more a result of overfitting the decision boundary. Even with representative training data, such an irregular decision boundary may not yield an unbiased model.*

##

Let's try and balance the above to solutions via cross-validation.  Using the `tune` function, pass in the training data, and a list of the following cost and $\gamma$ values: {0.1, 1, 10, 100, 1000} and {0.5, 1,2,3,4}.  Save the output of this function in a variable called `tune.out`.  

```{r}

set.seed(1)
tune.out <- tune(svm, y~., data = dat[train,], kernel = "radial", ranges = list(cost = c(0.1, 1, 10, 100, 1000), gamma = c(0.5, 1,2,3,4)))
summary(tune.out)

```

I will take `tune.out` and use the best model according to error rate to test on our data.  I will report a confusion matrix corresponding to the 100 predictions.  


```{r}
table(true=dat[-train,"y"], pred=predict(tune.out$best.model, newdata=dat[-train,]))
```

##

Comment on the confusion matrix.  How have we improved upon the model in question 2 and what qualifications are still necessary for this improved model.  

*We have improved our model a lot more and reduced overfitting. There is still some misclassification of class 1 as class 2 due to the underlying data imbalance in classes, though this isn't a fault of the model.*

# 
Let's turn now to decision trees.  

```{r}

library(kmed)
data(heart)
library(tree)

```

## 

The response variable is currently a categorical variable with four levels.  Convert heart disease into binary categorical variable.  Then, ensure that it is properly stored as a factor. 

```{r}

for (i in 1:length(heart$class)) {
  if (heart$class[i] > 0){
    heart$class[i] = 1
  }
}

heart$class = as.factor(heart$class)

```

## 

Train a classification tree on a 240 observation training subset (using the seed I have set for you).  Plot the tree.  

```{r}
set.seed(101)
train=sample(1:nrow(heart), 240)

tree.heart = tree(class~., heart, subset=train)
plot(tree.heart)
text(tree.heart, pretty=0)

```


## 

Use the trained model to classify the remaining testing points.  Create a confusion matrix to evaluate performance.  Report the classification error rate.  

```{r}

tree.pred = predict(tree.heart, heart[-train,], type="class")
with(heart[-train,], table(tree.pred, class))
1-(28+18)/57

```

##  

Above we have a fully grown (bushy) tree.  Now, cross validate it using the `cv.tree` command.  Specify cross validation to be done according to the misclassification rate.  Choose an ideal number of splits, and plot this tree.  Finally, use this pruned tree to test on the testing set.  Report a confusion matrix and the misclassification rate.  

```{r}
set.seed(101)
cv.heart = cv.tree(tree.heart, FUN = prune.misclass)
cv.heart

plot(cv.heart$size, cv.heart$dev, type = "b")
prune.heart = prune.misclass(tree.heart, best =3)
plot(prune.heart)
text(prune.heart, pretty=0)

tree.pred = predict(prune.heart, heart[-train,], type="class")
with(heart[-train,], table(tree.pred, class))
1-((26+17)/57)

```


##

Discuss the trade-off in accuracy and interpretability in pruning the above tree. 

*While the pruned tree misclassifies a little more than the non pruned tree, the pruned tree is also less complex and a lot more interpretative. Thal seems to be the most influential variable followed by ca and then cp. Understanding this is a lot more simpler with the pruned tree*

## 

Discuss the ways a decision tree could manifest algorithmic bias.  

*A decision tree can manifest algorithmic bias if it is trained on unrepresentative/biased data. This can lead the decision tree to enforce those biases when testing or predicting. Another algorithmic bias that can manifest is overfitting. This is when the decision tree is too closely trained on a subset of data and it picks up irrelevant patterns/noise for that specific subset and fails to generalize well to the larger data set. To avoid these issues, it is important to use representative data, randomize training/testing splits, and employ techniques like pruning if necessary*


